---
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# {walkast}

## The objective
R is very flexble to handle a language object for meta-programming porpose.
This package aims to provide a simple way to access a language object and 
you can manipulate it as you want by passing a visitor-function for printing, transforming, analyzing, and generating a new code.

## OOP or Functional Style?
It is natural for an OOP language to use visitor-pattern, and it is common for
a functional language to use pattern-matching. R, I think, is not suitable for
the use of those idea because semantics itself does not support those
functionalities.

## Related functions that already exist
* `codetools`::`walkCode` by [Luke Tierney](https://cran.r-project.org/web/packages/codetools/index.html)
* `pryr`::`call_tree` (NSE version is `pryr`::`ast`) by [Hadley Wickham](https://cran.r-project.org/web/packages/pryr/index.html)

see [vignette](./vignettes/related_ast_functions.md) for a comparison among these functions.


## Installation
```{r, eval = FALSE}
## install.packages("devtools")
devtools::install_github("tobcap/walkast")
library("walkast")
```

## Main functions
#### `walk_ast(expr, visitor)`

* expr: a language object (not permitted expression object)  
* visitor: S3 class that is an environment which must have six functions (leaf, call, hd, tl, first, last)

#### `make_visitor(leaf, call, hd, tl, first, last, vars)`
A helper which creates visitor-class.

* leaf: a function that manipulates leaf part of langauge object (a symbol or an atomic)
* call: a function that manipulates call part of langauge object (call object)
* hd: a function that manipulates caller part (head of call object)
* tl: a function that manipulates arguments part of call (tail of call object)
* first: a function that manipulates expr before running AST
* last: a function that manipulates expr after running AST

## Other helper functions
#### Printing
* `show_tree`
* `show_lisp`
* `show_r`

#### Replacing
* `replace`
* `nest_expr`

#### Conversion
* `to_list`
* `to_call`

#### Combination
* `%then%`

## Examples
```{r, message = FALSE}
library("walkast")
e1 <- quote(1+2)
```
```{r}
walk_ast(e1)
```

```{r}
walk_ast(e1, show_tree())
```

```{r}
mult2 <- make_visitor(leaf = function(x) if (is.numeric(x)) x * 2 else x)
walk_ast(e1, mult2)

add1 <- make_visitor(leaf = function(x) if (is.numeric(x)) x + 1 else x)
walk_ast(e1, add1)

walk_ast(e1, add1 %then% mult2)
walk_ast(e1, mult2 %then% add1)
```

```{r}
walk_ast(e1, replace(quote(`+`), quote(`-`)))
```

```{r}
walk_ast(e1, replace(2, quote(x)))
```

```{r}
e2 <- quote((1 + x) ^ 2)

nest_expr(e2, quote(x), 3)

nest_expr(e2, quote(1 + x), 3)

nest_expr(quote(1 + 1 / x), quote(x), 5)
```
