---
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# walkast

## The objective
This package aims to provide a simple way to access a language object and 
you can manipulate it as you want by passing a visitor-function for printing, transforming, analyzing, and generating a new code.

## OOP or Functional Style?
It is natural for an OOP language to use visitor-pattern, and it is common for
a functional language to use pattern-matching. R, I think, is not suitable for
the use of those ideas because semantics itself does not support those
functionalities.

## Related functions that already exist
* `codetools::walkCode` by [Luke Tierney](https://cran.r-project.org/web/packages/codetools/index.html)
* `pryr::call_tree` (NSE version is `pryr::ast`) by [Hadley Wickham](https://cran.r-project.org/web/packages/pryr/index.html)

see [vignette](./vignettes/related_ast_functions.md) for a comparison among these functions.


## Installation
```{r, eval = FALSE}
## install.packages("devtools")
devtools::install_github("tobcap/walkast")
library("walkast")
```

## Main functions

#### `walk_ast(expr, visitor)`

* expr: a language object (not accept an expression object)  
* visitor: an environment whose class name includes "visitor" and which must have six functions (leaf, call, hd, tl, first, last)

#### `make_visitor(leaf, call, hd, tl, initial, final, ...)`

A helper which creates visitor-class.

* `leaf()`: a function that manipulates leaf part of langauge object (a symbol or an atomic)
* `call()`: a function that manipulates call part of langauge object (call object)
* `hd()`: a function that manipulates caller part (head of call object)
* `tl()`: a function that manipulates arguments part of call (tail of call object)
* `initial()`: a function that manipulates expr before running AST
* `final()`: a function that manipulates expr after running AST
* `...`: arbitrary functions or variables that you want to use

    ```{r, message = FALSE, eval = FALSE}
    # you can use R6 class to make `visitor()` if you want
    v0 <- make_visitor(leaf = function(x) if(is.numeric(x)) x * 2 else x)
    
    # need to define all functions
    library(R6)
    v1 <- R6Class(
        "visitor"
      , public = list(
          leaf = function(x) if(is.numeric(x)) x * 2 else x
        , call = identity
        , hd = identity
        , tl = identity
        , initial = identity
        , final = identity
        )
      )$new()
    
    walk_ast(quote(1 + 2 * 3), v0)
    walk_ast(quote(1 + 2 * 3), v1)
    ```


## Other helper functions

#### Printing
* `show_tree()`
* `show_lisp(quote_bin = FALSE)`
* `show_r()`

#### Replacing
* `replace(before, after)`
* `nest_expr(expr, target, count)` this recursively calls walk_ast()

#### Conversion
* `to_list()`
* `to_call()`

#### Combination
* `%then%`

## Examples
```{r, message = FALSE}
library("walkast")
e1 <- quote(1 + 2)
```
```{r}
walk_ast(e1)
```

```{r}
walk_ast(e1, show_tree())
```

```{r}
mult2 <- make_visitor(leaf = function(x) if (is.numeric(x)) x * 2 else x)
walk_ast(e1, mult2)

add1 <- make_visitor(leaf = function(x) if (is.numeric(x)) x + 1 else x)
walk_ast(e1, add1)

walk_ast(e1, add1 %then% mult2)
walk_ast(e1, mult2 %then% add1)
```

```{r}
walk_ast(e1, replace(quote(`+`), quote(`-`)))
```

```{r}
walk_ast(e1, replace(2, quote(x)))
```

```{r}
e2 <- quote((1 + x) ^ 2)

nest_expr(e2, quote(x), 3)

nest_expr(e2, quote(1 + x), 3)

nest_expr(quote(1 + 1 / x), quote(x), 5)
```

```{r}
e3 <- quote({
    x <- 1
    ++x
    print(x)
})

plus_plus <- make_visitor(
  call = function(x) 
    if (length(x) == 2 && identical(x[[1]], quote(`+`)) &&
        length(x[[2]]) == 2 && identical(x[[2]][[1]], quote(`+`)) &&
        is.symbol(sym <- x[[2]][[2]]))
    base::call("<-", sym, base::call("+", sym, 1)) else x
)

walk_ast(e3, plus_plus)

plus_plus2 <- make_visitor(
  call = function(x) {
    syms <- all.names(x)
    if (length(syms) == 3 &&
        syms[1:2] == c("+", "+") &&
        is.symbol(sym <- x[[2]][[2]]))
    base::call("<-", sym, base::call("+", sym, 1)) else x
  }
)

walk_ast(e3, plus_plus2)

```
## ToDo (someday)
* Get information of Type of each Node
* Analyze context and identify an environment where a function is called
* Check if expression is reducable


## R's grammer
* see https://github.com/wch/r-source/blob/trunk/src/main/gram.y#L334-L454
```
%%
prog	:	END_OF_INPUT
	|	'\n'
	|	expr_or_assign '\n'
	|	expr_or_assign ';'
	|	error
	;
expr_or_assign  :    expr
                |    equal_assign
                ;
equal_assign    :    expr EQ_ASSIGN expr_or_assign
                ;
expr	: 	NUM_CONST
	|	STR_CONST
	|	NULL_CONST
	|	SYMBOL
	|	'{' exprlist '}'
	|	'(' expr_or_assign ')'
	|	'-' expr %prec UMINUS
	|	'+' expr %prec UMINUS
	|	'!' expr %prec UNOT
	|	'~' expr %prec TILDE
	|	'?' expr
	|	expr ':'  expr
	|	expr '+'  expr
	|	expr '-' expr
	|	expr '*' expr
	|	expr '/' expr
	|	expr '^' expr
	|	expr SPECIAL expr
	|	expr '%' expr
	|	expr '~' expr
	|	expr '?' expr
	|	expr LT expr
	|	expr LE expr
	|	expr EQ expr
	|	expr NE expr
	|	expr GE expr
	|	expr GT expr
	|	expr AND expr
	|	expr OR expr
	|	expr AND2 expr
	|	expr OR2 expr
	|	expr LEFT_ASSIGN expr
	|	expr RIGHT_ASSIGN expr
	|	FUNCTION '(' formlist ')' cr expr_or_assign %prec LOW
	|	expr '(' sublist ')'
	|	IF ifcond expr_or_assign
	|	IF ifcond expr_or_assign ELSE expr_or_assign
	|	FOR forcond expr_or_assign %prec FOR
	|	WHILE cond expr_or_assign
	|	REPEAT expr_or_assign
	|	expr LBB sublist ']' ']'
	|	expr '[' sublist ']'
	|	SYMBOL NS_GET SYMBOL
	|	SYMBOL NS_GET STR_CONST
	|	STR_CONST NS_GET SYMBOL
	|	STR_CONST NS_GET STR_CONST
	|	SYMBOL NS_GET_INT SYMBOL
	|	SYMBOL NS_GET_INT STR_CONST
	|	STR_CONST NS_GET_INT SYMBOL
	|	STR_CONST NS_GET_INT STR_CONST
	|	expr '$' SYMBOL
	|	expr '$' STR_CONST
	|	expr '@' SYMBOL
	|	expr '@' STR_CONST
	|	NEXT
	|	BREAK
	;
cond	:	'(' expr ')'
	;
ifcond	:	'(' expr ')'
	;
forcond :	'(' SYMBOL IN expr ')'
	;
exprlist:
	|	expr_or_assign
	|	exprlist ';' expr_or_assign
	|	exprlist ';'
	|	exprlist '\n' expr_or_assign
	|	exprlist '\n'
	;
sublist	:	sub
	|	sublist cr ',' sub
	;
sub	:
	|	expr
	|	SYMBOL EQ_ASSIGN
	|	SYMBOL EQ_ASSIGN expr
	|	STR_CONST EQ_ASSIGN
	|	STR_CONST EQ_ASSIGN expr
	|	NULL_CONST EQ_ASSIGN
	|	NULL_CONST EQ_ASSIGN expr
	;
formlist:
	|	SYMBOL
	|	SYMBOL EQ_ASSIGN expr
	|	formlist ',' SYMBOL
	|	formlist ',' SYMBOL EQ_ASSIGN expr	
	;
cr	:					{ EatLines = 1; }
	;
%%
```