---
title: "Walking AST of R-language"
author: "tobcap@github"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Walking AST of R-language}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Comparison of main features

|   | codetools | pryr | walkast |
|:-:|:-:|:-:|:-:|
| show ast | `walkCode(expr)` | `call_tree(expr)` | `walk_ast(expr)` |
| modify call | `walkCode(expr, replaceLeaf)`^[replaceLeaf <- function(before, after) makeCodeWalker(leaf = function(e, w) if (identical(e, before)) after else e, call = function(e, w) as.call(lapply(e, walkCode, w)))
] | `modify_call(call, new_args)`, `modify_lang(x, f, ...)` | `walk_ast(expr, replace(...))` |
| print lisp-like expression | `showTree(expr)` | NA | `walk_ast(expr, to_lisp())` |
| to list() structure | `walkCode(expr, toList)`^[toList <- makeCodeWalker(call = function(e, w) lapply(e, walkCode, w), leaf = function(e, w) e)] | NA | `walk_ast(expr, to_list())` |

## Showing AST
```{r, echo = TRUE}
library(pryr); library(codetools); library(walkast)
expr1 <- quote(sum(1, x + y, sin(z * w)))
```
```{r, echo = TRUE}
walkCode(expr1)

call_tree(expr1)

walk_ast(expr1)
```
